Analysing the three-dimensional structure of \em{P. falciparum}

### 2017-07-03 doc/manuscript/paper.tex
MDS approach

% Title of paper
\title{My very cool title}

% List of authors, with corresponding author marked by asterisk
\author{Nelle Varoquaux$^\ast$,\\[4pt]
% Author addresses
\textit{Department of Statistics, University of California, Berkeley \\
Berkeley Data Science Fellow}
\\[2pt]
% E-mail address for correspondence
{eaheron@tcd.ie}}

@@

Let us first introduce some notations. Hi-C data can be summarized as an $n
\times n$ symmetric contact count matrix, where each row and column corresponds to a
genomic window (or loci), and each entry $c_{ij}$ refers to the number of time
$i$ and $j$ have been seen interacting with one another. The matrix is hollow

Hi-C data suffers from technical and biological biases that need to be
corrected. \fixme{blha}

@@

\subsection{3D structure inference}

A crucial step of our analysis of the \textit{P. falciparum} genome relied on
visual introspection of a consensus 3D model of the genome. Being able to
quickly visualize how chromosome folded, where families of genes lied in the
3D space or whether regions where enriched helped us integrate with other
sources of data.

Briefly, we cast the problem of finding a consensus 3D structure $X \in R^{n
\times n}$ as an optimization problem, either as minimizing a cost function,
(often measuring the error between euclidean distance of $\mathbf{X}$ and a
series of ``wish distances'') or maximizing a likelihood. The optimization
problem can either be unconstrained or constrained to reflect some prior
knowledge on the structure (size and shape of the nucleas, distances between
loci, etc).

We added constraints to our model based on prior knowledge, such as the size
of the nucleus, or how far apart two adjacent beads should lie, though we
showed these constraints where well satisfied even when performing an
unconstrained optimized. In this section, I will review the inference method
we used in \citet{ay:three-dimensional} as well as two newer methods, based on
well-grounded statistical method.

\subsubsection{Multidimensional scaling}

\begin{equation}\label{eq:mds1} 
\renewcommand{\arraystretch}{2} 
\begin{array}{ccll} 
\underset{\mathbf{X}}{\text{minimize}} & & 
\underset{(i,j) \in \mathcal{D}}{\sum} w_{ij} \big(d_{ij}(\mathbf{X}) - \delta_{ij}\big)^2
\,,&\\ 
\end{array} 
\end{equation}

where $w_{ij}$ corresponds to weight.

@@

\subsubsection{Structure stability}
- Structure stability
- at least 100 iterations
- maybe compare results with chromSDE, PCA \& ShRec3D 
- 
\subsubsection{Volume exclusion models}

@@

\subsection{Simulation study application}

>>>>

% Title of paper
\title{Analysing the three-dimensional structure of \em{P. falciparum}}

% List of authors, with corresponding author marked by asterisk
\author{Nelle Varoquaux$^\ast$,\\[4pt]
% Author addresses
\textit{Department of Statistics, University of California, Berkeley \\
Berkeley Data Science Fellow}
\\[2pt]
% E-mail address for correspondence
{nelle@berkeley.edu}}

@@

@@

\subsection{Inferring the 3D structures}

An important part of analysing the 3D structure of the genome of \textit{P.
falciparum} relied on having consensus models for each stage of the life
cycle. The method presented in \citet{ay:three-dimensional} relied on an
approach called ``multidimensional scaling'' (MDS), widely used in the
literature for the purpose of 3D structure inference
\citep{duan:three-dimensional, varoquaux:statistical, bau:, tanizawa:}. Each chromosomes is
modeled a series of beads, spaced ~10~kb apart. MDS-based method consists of
two steps: first convert the
contact counts $c_{ij}$ into ``wish-distances'' $\delta_{ij}$; then attempt
to place each bead in 3D such that the euclidean distances $d_{ij}$ separating
beads $i$ and $j$ matches as closely as possible the wish distance
$\delta_{ij}$.

While these approaches are widely used, since better motivated and more
flexible models to infer 3D consensus structures from contact counts have been
proposed.

\subsubsection{Constrained multi-dimensional scaling}

The method presented in \citet{ay:three-dimensional} relied on an
approach called ``multidimensional scaling'' (MDS), widely used in the
literature for the purpose of 3D structure inference
\citep{duan:three-dimensional, varoquaux:statistical, bau:, tanizawa:}. Each chromosomes is
modeled a series of beads, spaced ~10~kb apart. MDS-based method consists of
two steps: first convert the
contact counts $c_{ij}$ into ``wish-distances'' $\delta_{ij}$; then attempt
to place each bead in 3D such that the euclidean distances $d_{ij}$ separating
beads $i$ and $j$ matches as closely as possible the wish distances
$\delta_{ij}$.

\paragraph{Wish distances}

A crucial step of MDS-based method rely on finding a good set of
``wish-distances'' from Hi-C contact count data.

\paragraph{Optimization}

Given the resulting wish-distances, we cast the inference of the consensus
structure $\mathbf{X}^{3 \times n}$ as the following constrained weighted MDS:

\begin{equation}\label{eq:mds1} 
\renewcommand{\arraystretch}{2} 
\begin{array}{ccll} 
\underset{\mathbf{X}}{\text{minimize}} & & 
\underset{(i,j) \in \mathcal{D}}{\sum} \frac{1}{\delta^2_{ij}}
\big(d_{ij}(\mathbf{X}) - \delta_{ij}\big)^2 & \\
\text{subject to} & & d_{ij} \leq r^2_\text{max} \quad \forall i, j &\,,\\
& & d_{i, i+1} \leq b_\text{max} \quad i = \{i:n|\text{chr}_i = \text{chr}_{i+1}\,,&\\ 
\end{array} 
\end{equation}

The various constraints corresponds to {\em prior} knowledge on the 3D
structure, and vary across organisms, time-points of the life cycle and
cell-type. For {\em P. falciparum}, the constraints are as follow.

\begin{itemize}
\item All beads lie in a spherical nucleus centered on the origin and of a
certain size.
\item Two adjacent beads must not be too far apart. Leveraging the work of
\citet{berger}, we assume that $b^\text{max} = 91~nm$
\end{itemize}

Note that in practice, the constraints are mostly intrinsic to the data and
thus can be dropped.

@@

\subsubsection{Structure stability}
- Structure stability
- at least 100 iterations
- maybe compare results with chromSDE, PCA \& ShRec3D 
-

\subsubsection{Volume exclusion models}

@@
